\documentclass[12pt]{article}
\usepackage[breaklinks=true]{hyperref}
\usepackage[margin=1in]{geometry}

\usepackage{color}

\definecolor{pblue}{rgb}{0.13,0.13,1}
\definecolor{pgreen}{rgb}{0,0.5,0}
\definecolor{pred}{rgb}{0.9,0,0}
\definecolor{pgrey}{rgb}{0.46,0.45,0.48}

\usepackage{listings}
\lstset{language=bash,
  showspaces=false,
  showtabs=false,
  breaklines=true,
  showstringspaces=false,
  breakatwhitespace=true,
  commentstyle=\color{pgreen},
  keywordstyle=\color{pblue},
  stringstyle=\color{pred},
  basicstyle=\ttfamily,
  frame=single,
  moredelim=[il][\textcolor{pgrey}]{$$},
  moredelim=[is][\textcolor{pgrey}]{\%\%}{\%\%}
}

\title{\textbf{Week 08} \\
\Large Exam, Cron, Curl, ASCII}
\author{
	Melvyn Ian Drag
}
\date{\today}


\begin{document}
\maketitle

\begin{abstract}
A `paella' of things to learn today. We will start with you demonstrating that you can configure a postgres server and do an interesting query on it. Then we will play with `curl' to make a variety of interesting http requests. Then we'll use `cron' to schedule tasks on our computer. Finally I'll set you up for what's to come later as we look at a text encoding called `ASCII'
\end{abstract}


\section{Exam}
Allocate time from 7:00 - 8:00 for the exam.

\section{Curl}
\subsection{Intro and Using Github to Test}
What is HTTP/HTTPS?

A prototcol used on the internet. You send some data to a server, the server responds with some data. As you will see with many data transmission technologies with computers, there is a header and a body. The header typically has meta data and bytes required by certain bits of computer hardware to process the body. Then the body contains the actual information.

HTTP/HTTP work with a few verbs. We'll just focus on two of them for now, but there are ( at least ) 4. 

Make some get, post requests. 

We're going to use this tutorial here:
\url{https://gist.github.com/joyrexus/85bf6b02979d8a7b0308}

Do one with authentication and a REST API.

\subsection{What kind of headers does curl take?}
What kind of header does it take?

Can we do two factor authentication with the REST API?

\subsection{}

\section{Cron}
8:45 - 9:30

Schedule a cowsay

Schedule a system backup

\section{Homework}
9:44 - 9:45 discuss homework

\end{document}
